# Part-to-whole

```{r, include = FALSE}
source("R/setup.R")
```

```{r, message=FALSE}
library(treemap)
library(faux)
library(tidyverse)
```

I've been wanting to get more familiar with treemap plots. 

## Simulate data

I always like to start my exploration with simulated data. The mixed design simulation functions below are from [faux](https://debruine.github.io/faux/articles/sim_mixed.html).

Make a school with 5 grades, 1-4 classes per grade, and 15-35 students per class.

```{r pth-data-sim}
set.seed(8675309) # for reproducible data simulation

school <- add_random(grade = 5) %>%
  add_random(class = sample(1:4, nrow(.), TRUE), 
             .nested_in = "grade") %>%
  add_random(student = sample(15:35, nrow(.), TRUE), 
             .nested_in = "class")

head(school)
```

## Basic Treemap

The basic version is OK, but I think we can do better here.

```{r, pth-treemap-basic}
school %>% 
  count(grade, class) %>%
  treemap(
    index = c("grade", "class"), 
    vSize = "n"
  )
```

## Adjust Labels

First, I'll adjust the label size and alignment. I adapted this code from a book that Emily Nordmann and I recently wrote: [Applied Data Skills](https://psyteachr.github.io/ads-v1/custom.html#treemap). For `fontsize.labels` and `align.labels`, I added vector names to help me remember which is which, but the names aren't used by `treemap()` at all, just the order.

The class labels aligned to the bottom right are too close to the edges, so give them some padding with `xmod.labels` and `ymod.labels`. I'm also going to remove the transparent grey label background with `bg.labels = 0`

```{r pth-treemap-labels}
school %>% 
  count(grade, class) %>%
  treemap(
    index = c("grade", "class"), 
    vSize = "n", 
    title = "",
    # set different label sizes for each type of label
    fontsize.labels = c(grade = 30, class = 15), 
    # remove shaded background
    bg.labels = 0,
    # set different alignments for two label types
    align.labels = list(
      grade = c("left", "top"), 
      class = c("right", "bottom")
    ),
    xmod.labels = c(0, -.1), # nudge class labels left
    ymod.labels = c(0, .1)   # nudge class labels up
  )
```

## Colour scheme

I'd like to use the #30DayChartChallenge theme colours for my plots, so let's set that up. I'm going to name the colours after the chapter they represent.

```{r pth-dcc-colours}
#| code-summary: Theme Colours
dcc_theme <- list(
  comp = "#4864DB", # ~royalblue
  dist = "#10D4CF", # ~darkturquoise
  rels = "#8DEA74", # ~lightgreen
  time = "#F5C748", # ~tan1
  cert = "#EA4E11"  # ~orangered2
)
```

```{r pth-treemap-colours}
#| code-summary: Plot
school %>% 
  count(grade, class) %>%
  treemap(
    index = c("grade", "class"), 
    vSize = "n", 
    title = "",
    palette = dcc_theme,
    fontsize.labels = c(grade = 30, class = 10), 
    bg.labels = 0,
    align.labels = list(
      grade = c("left", "top"), 
      class = c("center", "center")
    ),
    xmod.labels = c(0, -.1),
    ymod.labels = c(0, .1)
  )
```

## Change text

Now I want to upgrade the labels. It's probably easiest to do this in the data using `mutate()`. I'll set them to the kind of primary school grade labels we use in Scotland, where each grade is P1, P2, ... and each class within a grade gets a letter suffix. 

```{r school2}
#| code-summary: New labels
school2 <- school %>% 
  count(grade, class) %>%
  mutate(grade = gsub("grade", "P", grade)) %>%
  group_by(grade) %>%
  mutate(class = paste0(grade, letters[1:n()])) %>%
  ungroup() %>%
  mutate(class = paste0(class, "\nn = ", n))
```

Finally, make the borders thicker. It's starting to look like a Mondrian!

```{r pth-treemap-final}
#| fig-cap: Treemap plot of classes nested in grades (P1 to P5). There are 2-4 classes in each grade with 16-35 students. The size of each box represents the size of the class, and the colour of the box represents the grade.
day1_plot <- school2 %>%
  treemap(
    index = c("grade", "class"), 
    vSize = "n", 
    title = "",
    palette = dcc_theme,
    fontcolor.labels = "white",
    fontsize.labels = c(grade = 30, class = 15), 
    bg.labels = 0,
    align.labels = list(
      grade = c("left", "top"), 
      class = c("right", "bottom")
    ),
    xmod.labels = c(0, -.1),
    ymod.labels = c(0, .1),
    border.lwds = 5 # thicker borders
  )
```

## Saving a treemap

I tried `ggsave()` first and it took a *really* long time to save, and then the resulting image was blank. I should remember that not everything is ggplot; the output of the `treemap()` function is a list that draws a plot as a side effect (unless you set `draw = FALSE`), so you have to use base R plot saving techniques.

I couldn't figure out how to re-plot from the `day1_plot` object, so just had to retype the plot. This didn't work well in an Rmd code chunk (I think treemap can sense if it's in a code chunk with inline plot rendering and overrides the dev), so I had to run it in the console. 

```{r pth-treemap-save, eval = FALSE}
#| code-summary: Save treemap

# open the png file
ragg::agg_png("images/day1.png", 
              width = 8, height = 4.5, 
              background = "transparent",
              units = "in", res = 150)

# plot the treemap
school2 %>%
  treemap(
    index = c("grade", "class"), 
    vSize = "n", 
    title = "",
    palette = dcc_theme,
    fontcolor.labels = "white",
    fontsize.labels = c(grade = 30, class = 15), 
    bg.labels = 0,
    align.labels = list(
      grade = c("left", "top"), 
      class = c("right", "bottom")
    ),
    xmod.labels = c(0, -.1),
    ymod.labels = c(0, .1),
    border.lwds = 5 # thicker borders
  )

# close the png file
while (!is.null(dev.list())) dev.off()
 
```