# Mountains

```{r, include = FALSE}
source("R/setup.R")
```


Munros are the hills in Scotland over 3000 feet high. Many Scottish people are obsessed with [Munro bagging](https://www.visitscotland.com/see-do/active/walking/munro-bagging/) — trying to see how many of the 282 Munros you can climb. I've lived in Scotland since 2003, but I still haven't been up a Munro!

So I thought for the mountains prompt, I'd map all of the Munros in Scotland by height. Then someday, when I bag a munro, I can mark it on the map.

```{r}
#| code-summary: Setup

library(tidyverse)     # for data wrangling
library(sf)            # for maps
library(rnaturalearth) # for map coordinates
library(ggthemes)      # for the map theme
library(plotly)        # for interactive plots
library(showtext)      # for fonts

# install a good Scottish font
# https://www.fontspace.com/hill-house-font-f40002
font_add(family = "Hill House",
         regular = "fonts/Hill_House.otf")
showtext_auto()
```

## Data

[The Database of British and Irish Hills v17.3 ](http://www.hills-database.co.uk/downloads.html) has a table of the Munros, with columns for many years (I guess which hills are classified as Munros changes over time). Let's get just the current munros and fix some of the names.

```{r}
munros <- read_csv("data/munrotab_v8.0.1.csv",
                   show_col_types = FALSE) %>%
  filter(`2021` == "MUN") %>%
  select(-c(`1891`:`2021`)) %>% # get rid of the year columns
  rename(height_m = "Height (m)", 
         height_ft = "Height\n(ft)")
```

Make a quick histogram of their heights to get an overview of the data. I'd usually use the metric system, but since Munros are defined as hills over 3000 feet, I'll use feet.

```{r mountains-histogram}
ggplot(munros, aes(x = height_ft)) +
  geom_histogram(binwidth = 100, boundary = 0, 
                 color = "black", fill = "white")
```

## Map of Scotland

Now I need a map of Scotland.

```{r}
scotland_sf <- ne_states(geounit = "Scotland", 
                         returnclass = "sf")
```

The munro table locates the peaks using grid coordinates, and the map uses latitude and longitude. So I translated the grid coordinates to latitude and longitude using [Stackoverflow code from hrbrmstr](https://stackoverflow.com/questions/23017053/how-to-convert-uk-grid-reference-to-latitude-and-longitude-in-r).


```{r, echo = FALSE}
# takes numeric east/north generated from the os.grid.parse() function
# i shld have made it take the vector the os.grid.parse() returns but 
# we'll save that for a proper package version

os.grid.to.lat.lon <- function(E, N) {

  a <- 6377563.396
  b <- 6356256.909
  F0 <- 0.9996012717
  lat0 <- 49*pi/180
  lon0 <- -2*pi/180
  N0 <- -100000
  E0 <- 400000
  e2 <- 1 - (b^2)/(a^2)
  n <- (a-b)/(a+b)
  n2 <- n^2
  n3 <- n^3

  lat <- lat0
  M <- 0

  repeat {

    lat <- (N-N0-M)/(a*F0) + lat

    Ma <- (1 + n + (5/4)*n2 + (5/4)*n3) * (lat-lat0)
    Mb <- (3*n + 3*n*n + (21/8)*n3) * sin(lat-lat0) * cos(lat+lat0)
    Mc <- ((15/8)*n2 + (15/8)*n3) * sin(2*(lat-lat0)) * cos(2*(lat+lat0))
    Md <- (35/24)*n3 * sin(3*(lat-lat0)) * cos(3*(lat+lat0))
    M <- b * F0 * (Ma - Mb + Mc - Md)

    if (N-N0-M < 0.00001) { break }

  }

  cosLat <- cos(lat)
  sinLat <- sin(lat)

  nu <- a*F0/sqrt(1-e2*sinLat*sinLat)
  rho <- a*F0*(1-e2)/((1-e2*sinLat*sinLat)^1.5)

  eta2 <- nu/rho-1

  tanLat <- tan(lat)
  tan2lat <- tanLat*tanLat
  tan4lat <- tan2lat*tan2lat
  tan6lat <- tan4lat*tan2lat

  secLat <- 1/cosLat
  nu3 <- nu*nu*nu
  nu5 <- nu3*nu*nu
  nu7 <- nu5*nu*nu

  VII <- tanLat/(2*rho*nu)
  VIII <- tanLat/(24*rho*nu3)*(5+3*tan2lat+eta2-9*tan2lat*eta2)
  IX <- tanLat/(720*rho*nu5)*(61+90*tan2lat+45*tan4lat)
  X <- secLat/nu
  XI <- secLat/(6*nu3)*(nu/rho+2*tan2lat)
  XII <- secLat/(120*nu5)*(5+28*tan2lat+24*tan4lat)
  XIIA <- secLat/(5040*nu7)*(61+662*tan2lat+1320*tan4lat+720*tan6lat)

  dE <- (E-E0)
  dE2 <- dE*dE
  dE3 <- dE2*dE
  dE4 <- dE2*dE2
  dE5 <- dE3*dE2
  dE6 <- dE4*dE2
  dE7 <- dE5*dE2

  lon <- lon0 + X*dE - XI*dE3 + XII*dE5 - XIIA*dE7
  lat <- lat - VII*dE2 + VIII*dE4 - IX*dE6

  lat <- lat * 180/pi
  lon <- lon * 180/pi

  return(c(lat, lon))

}

```


```{r}
latlon <- map2(munros$xcoord, munros$ycoord, os.grid.to.lat.lon)

munros$lat <- map_dbl(latlon, `[[`, 1)
munros$lon <- map_dbl(latlon, `[[`, 2)
```

Then plot the latitude and longitude coordinates on the map, colored by height.

```{r mountains-map, fig.width = 5, fig.height = 8}
ggplot() + 
  geom_sf(data = scotland_sf,
          mapping = aes(),
          color = "black", 
          fill = "grey",
          size = .1) +
  coord_sf(xlim = c(-8, 0), ylim = c(54, 61)) +
  geom_point(aes(x = lon, y = lat, color = height_ft), munros) +
  theme_map()
```


## Make it prettier

There's no munros on the Northern Isles, so I've cropped them out of the map. I also made the colours better. I added a label to `geom_point()`, which will produce a warning that it isn't used, but you'll see why in the next step.

The version of Hill House I have doesn't have lowercase letters, so I'm using all uppercase for the title.

```{r mountains-pretty, fig.width = 4.5, fig.height = 6}
#| fig.cap: The location and height of Scotland's Munros.
#| fig.alt: A map of Scotland with the location of all the munros (hills over 3000 feet) mapped with their height shown by colour.

munro_plot <- ggplot() + 
  geom_sf(data = scotland_sf,
          mapping = aes(),
          color = "chocolate4", 
          fill = "tan",
          size = .1) +
  coord_sf(xlim = c(-7.4, -2), ylim = c(54.8, 58.5)) +
  geom_point(aes(label = Name, color = height_ft, y = lat, x = lon), 
             data = arrange(munros, height_ft),
             size = 0.5) +
  scale_color_viridis_c(name = "Height (ft)",
                        limits = c(3000, 4500)) +
  labs(x = NULL, y = NULL,
       title = "THE MUNROS OF SCOTLAND",
       caption = "Data from The Database of British and Irish Hills | Plot by @lisadebruine") +
  theme_map() +
  theme(legend.position = c(0, 0),
        legend.background = element_blank(),
        panel.background = element_rect(fill = "transparent", color = "transparent"),
        plot.background = element_rect(fill = "lightskyblue1", color = "transparent"),
        plot.title = element_text(family = "Hill House", size = 26, hjust = .5)
  )

munro_plot
```

```{r, include = FALSE, eval = FALSE}
ggsave("images/day8.png", plot = munro_plot, 
       width = 4.5, height = 6, device = png)
```

## Make it interactive

The `r pkg("plotly")` package makes it pretty easy to make a ggplot interactive. Select an are to get a closer look, or hover over an individual munro to see why we added the unused `label` argument to `geom_point()` above.

I had to add some css to get the hover bar to stop acting weird, but it's still not quite right.

```{verbatim}
.js-plotly-plot .plotly .modebar { position: relative; top: -2em; }
.plotly.plot-container { padding-top: 2em; background-color: #B0E1FE;}
```

<style>
.js-plotly-plot .plotly .modebar { position: relative; top: -2em; }
.plotly.plot-container { padding-top: 2em; background-color: #B0E1FE;}
</style>

```{r mountains-interactive, fig.width = 5, fig.height = 6}
plotly::ggplotly(
  p = munro_plot, 
  tooltip = c("label", "colour")
)
```
