# Experimental

```{r, include = FALSE}
source("R/setup.R")
```

Today I'm going to use faux to generate some simulated data for a factorial experimental design and make some plots.

```{r}
#| code-summary: Setup

library(faux)      # for data simulation
library(ggplot2)   # for plotting
library(patchwork) # for multi-panel plots
theme_set(theme_bw())
set.seed(8675309)  # for reproducibility
```

```{r, include = FALSE}
# reset to default
faux_options(plot = TRUE)
```


## Data

The `sim_design()` function is meant to make the simulation of data for factorial designs  easier. You can name the factors and their levels (see extensive [vignettes](https://debruine.github.io/faux/)), but here we'll just use the generic naming from faux and only specify the number of levels of each factor. By default, it simulates 100 observations in each between group, with means of 0 and standard deviations of 1. The function also produces a design plot.

### One factor

```{r exp-b2-sim}
b2 <- sim_design(between = 2, 
                 mu = c(0, 0.5))
```

For within-subject designs, you have to specify the correlation between each level. If there are any within factors, the `check_sim_stats()` table will include the sampled correlations. You can turn off the design plot with `plot = FALSE`. 

```{r}
w2 <- sim_design(within = 2, 
                 mu = c(0, 0),
                 sd = c(1, 5),
                 r = 0.5,
                 plot = FALSE)
check_sim_stats(w2)
```



### Two factors

You can also give the factors names and labels. You can set the correlations separately for each between-subject cell.

```{r}
# df version uses row and column labels, 
# so allows rows and columns to be out of order
mu <- data.frame(cat = c(0, 0.5),
                 dog = c(0.25, 1),
                 row.names = c("am", "pm"))

w2b2 <- sim_design(within = list(time = c("am", "pm")), 
                   between = list(pet = c("cat", "dog")),
                   mu = mu,
                   dv = list(y = "Something Score"),
                   r = list(cat = 0,
                            dog = .5),
                   long = TRUE)
check_sim_stats(w2b2)
```

## Plots

### Data Plots

Now it's time for the fun part: making plots! The `plot()` function makes it easy to plot data made with faux. See the [plotting vignette](https://debruine.github.io/faux/articles/plots.html) for more details.

The output is a ggplot, so you can add more ggplot options to it.

```{r exp-b2-plot}
plot(b2, geoms = c("violin", "box"), palette = "Accent") +
  ggtitle("Between-Subject Factor")
```

### Design Plots

You can also plot the design (not the data). If you plot a violin, the shape will be estimated by simulating a very large sample with these parameters.

```{r exp-w2-plot}
get_design(w2) %>%
  plot(geoms = c("violin", "pointrangeSD"), palette = "Dark2") +
  ggtitle("Within-Subject Factor")
```

### Combo Plots

Create a design plot for the main effects and interaction in `w2b2`. You can specify the order color, x, facet row(s), facet col(s) [this function needs a little work to be more intuitive]. Add a title and custom colours (you'll get a warning about the duplicate scale for fill; just ignore it).

```{r, warning=FALSE, message=FALSE}
geoms <- c("violin", "pointrangeSD")
b_colors <- scale_color_manual(values = c("red3", "dodgerblue3"), guide = "none")
w_colors <- scale_color_manual(values = c("darkorchid", "darkorange"), guide = "none")
b_fill <- scale_fill_manual(values = c("red3", "dodgerblue3"), guide = "none")
w_fill <- scale_fill_manual(values = c("darkorchid", "darkorange"), guide = "none")

plot_b2 <- get_design(w2b2) %>% 
  plot(geoms = geoms, "pet", "pet") + 
  b_colors + b_fill +
  labs(x = NULL, title = "Main Effect of Pet Type")

plot_w2 <- get_design(w2b2) %>% 
  plot(geoms = geoms, "time", "time") + 
  w_colors + w_fill +
  labs(x = NULL, title = "Main Effect of Time of Day")

plot_w2b2 <- get_design(w2b2) %>% 
  plot(geoms = geoms, "time", "pet") + 
  w_colors + w_fill +
  labs(x = NULL, title = "Interaction between Time of Day and Pet Type")

plot_b2w2 <- get_design(w2b2) %>% 
  plot(geoms = geoms, "pet", "time") + 
  b_colors + b_fill +
  labs(x = NULL, title = "Interaction between Pet Type and Time of Day")
```

Combine with `r pkg("patchwork")` 

```{r exp-day10, fig.width = 8, fig.height = 8}
#| fig.cap: Plots of the parameters for a simulated 2B2W experimental design.
#| fig.alt: A set of violin plots of the two main effects in a 2-between by 2-within design, plus two different views of the interaction plot.

layout <- paste("AB", "CC", "DD", sep = "\n")
plot_b2  + plot_w2 + plot_w2b2 + plot_b2w2 +
  plot_layout(design = layout)
```

```{r, include = FALSE, eval = FALSE}
ggsave("images/day10.png", width = 8, height = 8, device = png)
```
