# Colours {#colours}

```{r, include = FALSE}
source("R/setup.R")
library(dplyr)
```

Hover over a colour to see its R name.

<style>
  .colordemo { list-item-style: none; padding: 0; margin: 0;}
  
  .colordemo li, main ul.colordemo > li:first-child { 
    display: inline-block; 
    border: 0.25px solid black; 
    margin: 0px; 
    padding: 1.3em 0.2em 0.2em 1.2em; 
  }
  
  .colordemo li span { 
    display: none; 
    background-color: white; 
    border: 0.25px solid black; 
    padding: 0.25em 0.5em; 
    border-radius: 0 1em 1em 1em;
  }
  .colordemo li:hover span { 
    display: inline; 
    position: absolute;
  }
</style>

```{r colour-demo, echo = FALSE, results='asis'}
name <- colours(TRUE)
rgb <- col2rgb(name) 
hsv <- rgb2hsv(rgb)
hex <- as.hexmode(rgb) %>% as.character() %>% 
  apply(2, paste, collapse = "")

colors <- tibble(name = name,
       hue = hsv[1, ] %>% round(2),
       sat = hsv[2, ] %>% round(2),
       val = hsv[3, ] %>% round(2),
       hex = hex) %>%
  arrange(hue, sat, val)

glue::glue("    <li style='background-color: #{colors$hex};'><span>{colors$name}</span></li>") %>%
  paste(collapse = "\n") %>%
  paste0("\n<ul class='colordemo'>\n", ., "\n</ul>\n") %>%
  cat()
```

